---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
``` {r data reading}
activity <- read.csv(unz("activity.zip", "activity.csv"), stringsAsFactors=FALSE, na.strings = "NA")
activity$date <- as.Date(activity$date, format= "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
``` {r mean and median}
hist(aggregate(steps ~ date, data = activity, sum)$steps, xlab = "Daily steps", main = "Histogram of Daily Steps")
options(scipen=2)
StepsMean <- mean(aggregate(steps ~ date, data = activity, sum)$steps)
StepsMedian <- median(aggregate(steps ~ date, data = activity, sum)$steps)
```
The mean of the total number of steps taken by day is `r StepsMean`.

The median of the total number of steps taken by day is `r StepsMedian`.

## What is the average daily activity pattern?
``` {r average daily}
library(stringr)
activitymean <- aggregate(steps ~ interval, FUN = mean, data = activity)
activitymean$interval <- as.character(activitymean$interval)
activitymean$interval <- str_sub(paste("000", activitymean[,1], sep = ""), -4)
#activitymean$interval <- format(strptime(activitymean$interval, format="%H%M"), format = "%H:%M")
plot(activitymean$interval, activitymean$steps, type = "l", xaxt = "n", main = "Average Daily Activity pattern", xlab = "Intervals", ylab = "Steps average")
axis(1, xaxp=c("0000", "2400", 24), las=2)
intervalmaxsteps <- format(strptime(activitymean[which.max(activitymean$steps),1], format="%H%M"), format = "%H:%M")
```
On average, the interval with the maximum number of steps is `r intervalmaxsteps`.

## Imputing missing values
``` {r imputing missing values}
numberofnas <- nrow(activity[(is.na(activity$steps) | is.na(activity$date) | is.na(activity$interval)),])
## Creating the data set activitywonas with the Activity contents
activitywonas <- activity
## Populating all the missing values with the mean for 5-minute interval calculated previously
activitywonas[is.na(activity$steps),1] <- activitymean$steps
hist(aggregate(steps ~ date, data = activitywonas, sum)$steps, xlab = "Daily steps", main = "Histogram of Daily Steps")
options(scipen=2)
StepsMeanwoNAs <- mean(aggregate(steps ~ date, data = activitywonas, sum)$steps)
StepsMedianwoNAs <- median(aggregate(steps ~ date, data = activitywonas, sum)$steps)
if (StepsMean != StepsMeanwoNAs) {
  diffMean <- StepsMean - StepsMeanwoNAs
  messageMean <- paste("The missing values imputation impacted the mean calculation in ", diffMean, sep = "")
} else {
  messageMean <- "The missing values imputation did not impact the mean calculation."
}
if (StepsMedian != StepsMedianwoNAs) {
  diffMedian <- StepsMedian - StepsMedianwoNAs
  messageMedian <- paste("The missing values imputation impacted the median calculation in ", diffMedian, sep = "")
} else {
  messageMedian <- "The missing values imputation did not impact the median calculation."
}
```

The number of missing values in the data set is `r numberofnas`.

The mean of the total number of steps taken by day imputing the missing values is `r StepsMeanwoNAs`.

The median of the total number of steps taken by day imputing the missing values is `r StepsMedianwoNAs`.

`r messageMean`

`r messageMedian`

``` {r impact of imputing missing data}
library(reshape2)
library(ggplot2)
activitysum <- aggregate(steps ~ date, data = activity, sum, na.action = NULL)
activitysum[, "stepswonas"] <- aggregate(steps ~ date, data = activitywonas, sum)$steps
activitysum.long <- melt(activitysum, id.vars="date")
ggplot(activitysum.long, aes(date, value, fill=variable))+geom_bar(stat="identity",position="dodge") + scale_fill_discrete(name="date", labels=c("W/O Missing Values Imputation", "With Missing Values Imputation")) + labs(x = "Dates", y = "Steps Average")
```

## Are there differences in activity patterns between weekdays and weekends?
